---
title: "Overview"
format: html
editor_options: 
  chunk_output_type: console
---

This portion of the project will go here:
'Lastly, we’ll write up a document via quarto (just like we did with the homework) to describe our thought process and give examples of using the functions.'

first we should load some of our packages from the tidyverse package
```{r}
library(readr) #must install first
library(tidyverse) #must install first 
library(httr) #must install first
library(tidycensus) #must install first)
library(jsonlite) #must install first 
```

save the url as a string object
```{r}
URL_ids <- "https://api.census.gov/data/2023/acs/acs1/pums?get=SEX,PWGTP,MAR&SCHL=24"
#next we will load the data from the API website so that we can use it for our project
id_info <- httr::GET(URL_ids)
str(id_info, max.level = 1)
```

extracting the content portion because this is the data we will need to work with 
```{r}
parsed_pums_info <- fromJSON(rawToChar(id_info$content))
col_names <- parsed_pums_info[1, ]
data_rows <- parsed_pums_info[-1, ]
```

creating a tibble to view data more clearly
```{r}
pums_info <- as_tibble(data_rows)
colnames(pums_info) <- col_names 
```

displaying tibble
```{r}
pums_info
```

